---
title: "R tutorial"
author: "Jing Lyu"
date: "1/11/2022"
output:
  html_document: default
  pdf_document: default
---

# Installation
1. Install R
+ Go to r-project.org;
+ On the left side of the page, click the "CRAN" link;
+ Scroll down to the USA section, and click a link (say, Duke University);
+ Click the download link for your operating system (Windows, Mac, or Linux);
+ Follow the instructions to download the latest version of R.
2. Install Rstudio
+ Must install R first.
+ Go to www.rstudio.com and follow the instructions.

# Compile Rmd file into html

* click 'Knit': Knit to HTML.
* install rmarkdown package in R: install.packages('rmarkdown'), if you can not knit.


# Using R

## Getting instruction of a function

Question mark + the name of a function
```{r}

```

```{r}
# the instruction of 'lm' function
?lm
```

## Install a pacakge and library it

```{r,eval=F,include=,}
# eval=F: knitr will not run the code in the code chunk.
install.packages(package_name)
library(package_name)
```


## Read or write a dataset

Choose the right function depending on the type of data. Always check the correctness of your dataset after read it, i.e. num of rows and cols, type of each column (numeric or categorical).
```{r}
### for '.txt' data
## create a dataframe
dat = data.frame(X1= c(1,2,3),X2=c(10,20,30))
# write 'dat' into the txt type
write.table(dat,file = '/Users/jinglyu/Desktop/dat.txt')
# read '.txt' dataset
dattxt= read.table('/Users/jinglyu/Desktop/dat.txt')
dattxt
dim(dattxt)
### for '.csv' data
write.csv(dat,file = '/Users/jinglyu/Desktop/dat.csv')
datcsv = read.csv('/Users/jinglyu/Desktop/dat.csv')
```


## Manipulation on dataframe

```{r}
## check whether an object is a dataframe
is.data.frame(datcsv)
class(datcsv)
datcsv
## Slices
datcsv$X1
datcsv[,2:3]
datcsv[,-1]
## combine two dataframe
#by column
cbind(datcsv$X1,datcsv$X2)
#by row, the name of column must match
rbind(datcsv[,2:3],datcsv[,2:3])
# cbind and rbind are also workable for matrix

## get column names of a dataframe
colnames(datcsv)
```


## Manipulation on matrix and vector

```{r}
## create a matrix
mat1 = matrix(1:9,nrow = 3,ncol=3)
mat1
## convert a matrix into dataframe
dat2 = as.data.frame(mat1)
class(mat1)
class(dat2)
## convert a dataframe into matrix
mat1 = as.matrix(dat2)
## matrix slices
mat1[1:2,1:2]

vec1 = 1:3
## matrix transpose
t(mat1)
## extract diagonal of a matrix
diag(mat1)
## dimensions of a matrix
dim(mat1)
## length of a vector
length(vec1)
## sum of a vector
Y = seq(1,100,by=2)
sum(Y)
## product of all elements in a vector
prod(Y)
## mean of a vector
mean(Y)
## median of a vector
median(Y)
## variance and sd of vector
var(Y)
sd(Y)
summary(Y)
```

## For loop and if else

```{r}
Y = seq(1,100,by=2)
sum_Y=0
for(i in 1:length(Y))
{
  sum_Y = sum_Y+Y[i]
}
sum_Y
if(sum_Y>100)
{
  print('summation is bigger than 100')
}

mat1
## For loop by columns or rows for matrix
## sum of each column
apply(mat1,MARGIN = 2,FUN = sum)
## variance of each row
apply(mat1,MARGIN = 1, FUN = var)
```

## Define a function in R 
```{r}
nameoffunction = function(parameter1,parameter2)
{
  s=0
  if(parameter1>parameter2)
  {
    s=1
  }
  return(s)
}

nameoffunction(parameter1 = 2,parameter2 = 1)
```


## Plot in R

```{r}
X = seq(1,10,length.out = 50)
Y = sin(X)+rnorm(50)
Y2 = X/10
plot(X,Y,type = 'b',main='name of a plot',col='red') # b:Both (points and line)
lines(X,Y2,col='blue',type='b',pch=4)
abline(a=0.1,b=0.3,col='green')

## histogram
hist(Y)

## create a 2 by 2 panel for 4 plots
par(mfrow=c(2,2),mgp=c(1.8,.5,0), mar=c(3,3,1,1), oma=c(0,0,3,0), pty='m')
# for the meaning of these arguments, check in ?par
plot(X,Y,type = 'b',main='name of a plot',col='red')
hist(Y)
boxplot(Y,main = 'Boxplot')
```
## ggplot in R

```{r,eval=F}
install.packages('ggplot2')
```

```{r,warning=F,message=F}
library(ggplot2)
dat1 = data.frame(x=X, y=Y)
ggplot(data = dat1, aes(x = x, y = y)) +
  geom_line(linetype = "dashed")+
  geom_point() + labs(title = 'name of a plot')

# histogram
ggplot(dat1, aes(x=y))+
  geom_histogram(color="black", fill="white") +
  geom_vline(data=dat1, aes(xintercept=mean(Y), color="red"),
             linetype="dashed") + labs(title = 'name of a plot')
```

## Data manipulation with dplyr

+ A powerful R-package to manipulate, clean and summarize unstructured data. 
+ It makes data exploration and data manipulation easy and fast in R.

```{r,eval=F}
install.packages("tidyverse")
```

```{r,warning=F,message=F}
# Create DataFrame
library(dplyr)
df <- data.frame(
  id = c(10,11,12,13,14,15,16,17),
  name = c('sai','ram','deepika','sahithi','kumar','scott','Don','Lin'),
  gender = c('M','M','F','F','M','M','M','F'),
  dob = as.Date(c('1990-10-02','1981-3-24','1987-6-14','1985-8-16',
                  '1995-03-02','1991-6-21','1986-3-24','1990-8-26')),
  state = c('CA','NY',NA,NA,'DC','DW','AZ','PH'),
  scores = c(87, 68, 92, NA, 62, 77, NA, 82)
)
df
# select() multiple columns
df %>% select(c('id','name'))
# filter() by column Value
df %>% filter(gender == 'M')
# filter() by multiple conditions
df %>% filter(gender == 'M' & scores > 80)
# summarise on groupped data.
df %>% group_by(gender) %>% summarise(Mean.scores=mean(scores, na.rm=T)) %>% arrange(Mean.scores)
```



