---
title: "Discussion5"
author: "Jing Lyu"
date: "2/8/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Two-way ANOVA

In this section, we’ll use the built-in R data set 'ToothGrowth'. It includes information from a study on the effects of vitamin C on tooth growth in Guinea pigs.

The trial used 60 pigs who were given one of three vitamin C dose levels (0.5, 1, or 2 mg/day) via one of two administration routes: orange juice (OJ) or ascorbic acid (VC).

#### Visualization

```{r tooth, message=F, warning=F}
library(dplyr)
dat = ToothGrowth
str(dat)
table(dat$dose)
```
R treats 'dose' as a numeric variable based on the output. We’ll transform it to a factor variable.

```{r trans}
dat$dose <- factor(dat$dose,
                  levels = c(0.5, 1, 2),
                  labels = c("D0.5", "D1", "D2"))
table(dat$supp,dat$dose)
```

We have a well-balanced design.

To visualize the data grouped by the levels of the two factors, we can use a box plot.

```{r boxplot}
boxplot(len ~ supp * dose, data=dat, frame = FALSE,
        col = c("#00AFBB", "#E7B800"), ylab="Tooth Length")
```

#### Modeling

```{r aov}
model1 = aov(len ~ supp + dose, data = dat)
summary(model1)
# If you think these two variables have interactive effect:
model2 <- aov(len ~ supp * dose, data = dat)
summary(model2)
```

### Practice Project

#### Import dataset

* Harvard dataverse

```{r Harvard,message=F,warning=F}
#install.packages("haven")
#install.packages("tzdb")
library(haven) # readin sav file
star = read_sav("/Users/jinglyu/Desktop/ucd/2023W/STA207/CourseMaterials/Projects/STAR_Students.sav")%>%as.data.frame()
length(names(star)) # 379 variables
# check missing values
star.dat = star%>%dplyr::select(g1tmathss, g1classtype, g1schid, g1tchid)%>%mutate(na.count=rowSums(is.na(.))) 
table(star.dat$na.count)
star.temp1 = star.dat%>%filter(na.count==1)
head(star.temp1) # only missing math scores
star.dat = star.dat%>%na.omit()%>%dplyr::select(-na.count) # drop rows with missing data
head(star.dat)
str(star.dat)
star.dat$g1classtype = as.factor(star.dat$g1classtype)
star.dat$g1schid = as.factor(star.dat$g1schid)
star.dat$g1tchid = as.factor(star.dat$g1tchid)
# summarize mean scores by class types, teachers and schools
star.dat1 = star.dat%>%group_by(g1classtype,g1schid,g1tchid)%>%
  dplyr::summarise(math.mean=mean(g1tmathss))
head(star.dat1)
```

* AER package

```{r AER, message=F, warning=F}
#install.packages('AER')
library(AER)
library(dplyr)
data("STAR")
sapply(STAR,class)
dim(STAR)
# only keep 1st grade
STAR.dat = STAR%>%dplyr::select(math1, school1, experience1, tethnicity1, schoolid1, star1)%>% 
  na.omit() # remove rows with NA value in any column
str(STAR.dat)
```
